---
title: "Statistical Trending on Medical Device Surveillance Data"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Medical Device Surveillance Statistical Trending}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
load(file="../data/mds_ts.rda")
library(mdsstat)
```

The `mdsstat` package:

- Standardizes the output of various statistical trending algorithms
- Allows running of multiple algorithms on the same data
- Allows running of both **disproportionality** and **quality control** algorithms
- Creates lightweight analysis definitions and output files for auditability, documentation, and reproducibility

**Why?**

There are many ways to trend medical device event data. Some are drawn from the [quality control](https://en.wikipedia.org/wiki/Quality_control) discipline, others from disproportionality analysis used in pharmacoepidemiology, and yet others from the general field of statistics.

There is a need to rigorously compare and contrast these various methods to more fully understand their respective performance and applicability in surveillance of medical devices.

**How?**

The `mdsstat` package aims to provide a collection of statistical trending algorithms used in medical device surveillance. Furthermore, each algorithm is written with a standardized, reusable framework philosophy. The same input data can be fed through multiple algorithms. All algorithms return **results that can be sorted, stacked, and compared.**

This package is written in tandem with the `mds` package. These are complementary in the sense that:

- `mds` standardizes medical device event data.
- `mdsstat` standardizes the statistical trending of medical device event data.

While `mdsstat` algorithms can run on generic R data frames, additional efficiency and traceability benefits are derived by running on data frames generated by `mds::time_series()` from the `mds` package.

**Purpose of This Vignette**

- Introduce [the algorithms](#algos)
- Show how to [run one algorithm](#onealgo)
- Show how to run [multiple algorithms](#mutlialgo)

## Data: MAUDE Time Series Generated from `mds::time_series()`

The following examples use a sample list of three time series generated by `mds::time_series()` from the `mds` package, saved as `mds_ts` in this package. The underlying data were queried from the [FDA MAUDE API](https://open.fda.gov). Furthermore, a simulated exposure dataset was generated to provide exposure data.

```{r}
library(mdsstat)
data(mds_ts)
```


## The Algorithms {#algos}

This is the current list of algorithms available:

Function            Description
-----------------   ------------------------------------------------------------
`xbar()`            Shewhart x-bar Control Chart with 4 Western Electric Rules
`cusum()`           Cumulative Sum Control Chart with 4 Western Electric Rules
`ewma()`            Exponentially Weighted Moving Average
`sprt()`            Sequential Probability Ratio Test
`prr()`             Proportional Reporting Ratio
`ror()`             Reporting Odds Ratio
`gps()`             Gamma Poisson Shrinker (containing EBGM and EB05)
`bcpnn()`           Bayesian Confidence Propagation Neural Network
`lrt()`             Likelihoood-Ratio Test
`cp_mean()`         Mean-Shift Changepoint
`poisson_rare()`    Generic Poisson Test

These are planned/proposed algorithms to add:

Function            Description
-----------------   ------------------------------------------------------------
`chi_square()`      Chi-Square Test
`changepoint()`     Additional changepoint variants
`cox_stuart()`      Cox-Stuart Test
`uptrend()`         Linear Uptrend by Linear Modeling


## Run One Algorithm {#onealgo}

In basic usage, running an `mdsstat` algorithm requires two considerations:

1. Input data format (may be reused in other algorithms)
1. Algorithm parameter settings (unique to the algorithm)

Here are some example algorithm calls:

```
# Example mds_ts data
data <- mds_ts[[3]]
data$rate <- ifelse(is.na(data$nA), 0, data$nA) / data$exposure

# Four different algorithm calls
xbar(data)
prr(data)
xbar(data, ts_event=c(Rate="rate"), we_rule=2)
poisson_rare(data, p_rate=0.3)
```

### Input Data Format

Input data shall be either a generic data frame (general usage) or an `mds_ts` data frame. Both are conceptually structured like time series.

#### `mds_ts` Usage

`mds_ts` data frames are generated by `mds::time_series()` from the `mds` package. These data frames are already structured for seamless integration into `mdsstat` algorithms. 

Note the following:

- Disproportionality algorithms will run only if the `mds_ts` data contains the columns `nA`, `nB`, `nC`, and `nD`. These are generated by specifying device and event hierarchies using `mds` package functions.
- Algorithms run by default using the `nA` column for event occurrence. 
- If running on event rate is desired, you may calculate an additional field and specify that field using the `ts_event` parameter.

*Running an algorithm on event rate instead of event count*

```
data <- mds_ts[[3]]
data$rate <- ifelse(is.na(data$nA), 0, data$nA) / data$exposure
xbar(data, ts_event=c("Rate"="rate"))
```

#### General Usage: Count or Rate Data

A generic data frame contains two columns, `time` and `event`, where for each unique sequential time (numeric or Date), there corresponds a number indicating the event occurrence. The event occurrence may commonly be the count of events or event rate.

An example:

```
data <- data.frame(time=c(1:25), event=as.integer(stats::rnorm(25, 100, 25)))
xbar(data)
```

#### General Usage: Data for Disproportionality Analysis (DPA)

Because disproportionality analysis is run on count data in a 2x2 contingency table, this data frame requires five columns, `time`, `nA`, `nB`, `nC`, and `nD`. For each unique sequential time (numeric or Date), there corresponds a set of numbers indicating the event counts. The latter four columns correspond to counts for cells A through D of the contingency table.

An example:

```
data <- data.frame(time=c(1:25),
                   nA=as.integer(stats::rnorm(25, 25, 5)),
                   nB=as.integer(stats::rnorm(25, 50, 5)),
                   nC=as.integer(stats::rnorm(25, 100, 25)),
                   nD=as.integer(stats::rnorm(25, 200, 25)))
prr(data)
```

#### General Usage: Count/Rate and DPA Data

To run algorihtms on both counts/rates and DPA, just include all the above columns, such as:

```
data <- data.frame(time=c(1:25),
                   event=as.integer(stats::rnorm(25, 100, 25)),
                   nA=as.integer(stats::rnorm(25, 25, 5)),
                   nB=as.integer(stats::rnorm(25, 50, 5)),
                   nC=as.integer(stats::rnorm(25, 100, 25)),
                   nD=as.integer(stats::rnorm(25, 200, 25)))
xbar(data)
prr(data)
```

## Run Multiple Algorithms {#mutlialgo}

`mdsstat` makes it easy to run multiple algorithms and variants of the same algorithm on your data. 

Just two steps are required:

1. Use `define_algos()` to set a list of algorithms with corresponding parameter settings.
1. Use `run_algos()` to run the algorithms defined in Step 1 on your data.

For example:

```
# Your data
data <- mds_ts[[3]]
data$rate <- ifelse(is.na(data$nA), 0, data$nA) / data$exposure

# Save a list of algorithms to run
x <- list(prr=list(),
          xbar=list(),
          xbar=list(ts_event=c(Rate="rate"), we_rule=2),
          poisson_rare=list(p_rate=0.3))
algos <- define_algos(x)

# Run algorithms
run_algos(data, algos)
```

By default, `run_algos()` returns the results of each algorithm as a row in a data frame. This allows for easy tabular review of algorithm performance.

## One Algorithm Returned as a Data Frame Row

Similar to the default output of `run_algos()`, you may convert the output of any `mdsstat` algorithm from the default list to a data frame row. Simply use `test_as_row()` on any algorithm output.

For example:

```
data <- data.frame(time=c(1:25), event=as.integer(stats::rnorm(25, 100, 25)))
result <- xbar(data)
test_as_row(result)
```
